\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-01-31 fluB.tex
Manuscript revisions.

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages on the basis of haemagglutinin inhibition (HI) assays. 
Frequent reassortment between the segments of these two lineages has been noted in the past \cite{lindstrom1999}, but the effects of these reassortments have not been investigated in much detail.

@@

\section*{Introduction}
Seasonal influenza causes an estimated 250,000 to 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality.
Although the known antigenic diversity of influenza B viruses is less than that of influenza A viruses, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

Like all members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of seasonal human and non-human influenza A strains. 
Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013}), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.

Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/02/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}. 
These two lineages possess antigenically distinct HA surface glycoproteins allowing them to co-circulate in the human population.
All other influenza B segments also show a split into two lineages around that time.

Previously, through phylogenetic analyses the evolutionary rate, selective pressures and reassortment history of major clades of influenza B have been analyzed \cite{chen2008}, showing extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within Vic and Yam lineages.
Here, we expand upon the methods used previously and introduce another approach that reveals a evolutionarily intriguing pattern of reassortment in influenza B.
In this approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
By modelling these labels on a phylogenetic tree, reassortment events which result in the replacement of one segment lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
We use this method to reconstruct major reassortment events in the recent history of influenza B viruses and to partition diversity across segments of influenza B virus strains to quantify reassortment dynamics over time.

We show that despite extensive reassortment, PB1, PB2 and HA segments of influenza B viruses still survive as distinct Victoria (Vic) and Yamagata (Yam) lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
In other segments (PA, NP, NA, MP and NS) a single lineage has been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
This has occured through repeated reassortments and subsequent fixation of reassortant viruses within the influenza B population.

\begin{figure}[h]
 \centering		
	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
	\caption{\textbf{Schematic depiction of method.}
A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages.
B) We then transfer lineage labels from one tree to the same tips in another tree.
Label switches thus indicate reassortment events that combine lineages falling on different sides of the bifurcation specified earlier.
C) In the example above tip number 6 is determined to be such a reassortant.}
	\label{methodFig}
\end{figure}

@@

We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information). 
The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment and identifying whether the isolate was more closely related to B/Victoria/02/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/02/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.

Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/02/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).

\subsection*{Building trees}
Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate HKY models for each of the 3 codon partitions, and additionally used a recently described method for reporting estimates of realized synonymous and non-synonymous substitutions \cite{obrien2009}.
We used a flexible demographic model (Bayesian skyride \cite{minin2008}).
We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
We ran 3 independent MCMC chains, each with 200 million states, sampled every 20000 steps and discarded the first 10\% of the MCMC states as burn-in.
After assessing convergence of all 3 MCMC chains by Tracer, we combined all analyses to give a total of 27000 samples from the posterior distribution of trees.

Every sequence was assigned 7 discrete traits corresponding to the lineages of all other segments with which a strain was isolated (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
We inferred the ancestral state of lineages in each segment by using discrete traits (asymmetric CTMC matrix with BSSVS \cite{lemey2009}). Because the posterior set of trees have branches labelled with the inferred lineage in all other segments, we can detect inter-lineage reassortments whenever a trait transition is observed (i.e. Yam to Vic or Vic to Yam, see Figure \ref{methodFig}) between a node and its descendant in any segment. 
In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.

@@

\subsection*{Confirming findings with a larger dataset through lineage assignment}
The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/02/87 or B/Yamagata/16/88 sequences.
Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.

>>>>

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
\tbc{Expand abstract to broadly outline results and give a 1 sentence conclusion.}

@@

\section*{Introduction}
Seasonal influenza causes between 250,000 and 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}. 
Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments (a process known as reassortment). 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains and with non-human influenza A strains. 
Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2013}.

Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}. 
These two lineages now possess antigenically distinct HA surface glycoproteins \tbc{CITE} allowing them to co-circulate in the human population.
Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of major clades of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses between and within the Vic and Yam lineages \cite{chen2008}.
Here, we extend previous methods to reveal an evolutionarily intriguing pattern of reassortment in influenza B.
In our approach, membership to either the Victoria or Yamagata lineage in the tree of one segment is used to label the individual viruses in the tree of the other segments.
By modelling the transition between labels on a phylogenetic tree, reassortment events which result in the replacement of one segment's lineage by another show up as label changes along a branch (Figure \ref{methodFig}).
We use this method to reconstruct major reassortment events and quantify reassortment dynamics over time in influenza B viruses.

We show that despite extensive reassortment, three of the eight segments, PB1, PB2 and HA, still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
This has occured through repeated reassortments and subsequent fixation of reassortant viruses within the influenza B population.

\begin{figure}[h]
 \centering		
	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
	\caption{\textbf{Schematic analysis of reassortment patterns.}
	A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages, in this case, the Victoria and Yamagata bifurcation that occurred in the early 1980s.
	B) We then transfer lineage labels from one tree to the same tips in another tree.
	Label switches thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation.
	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant.
	\tbc{Relabel as "Segment 1" and "Segment 2" instead of "Tree 1" and "Tree 2".}
	}
	\label{methodFig}
\end{figure}

@@

We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in supplementary information). 
The segments of each strain were assigned to a lineage (either Vic or Yam) by making maximum likelihood trees of each segment \tbc{using what software?} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.

Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees (for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y).

\subsection*{Building trees}
Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
We used a flexible Bayesian skyride demographic model \cite{minin2008}.
We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.
We ran 3 independent MCMC chains, each with 200 million states, sampled every 20,000 steps and discarded the first 10\% of the MCMC states as burn-in.
After assessing convergence of all 3 MCMC chains by visual inspection using Tracer \tbc{CITE}, we combined samples across chains to give a total of 27,000 samples from the posterior distribution of trees.
\tbc{The MCC tree for each segment should be shown in supplemental information.  I would imagine one figure with eight panels.}

Every sequence was assigned 7 discrete traits corresponding to the lineages of all other segments with which a strain was isolated (e.g. PB1 tree had PB2, PA, HA, NP, NA, MP and NS as traits and V or Y as trait values).
We inferred the ancestral state of lineages in each segment by modelling transitions between these discrete traits using an asymmetric transition matrix \tbc{CITE asymmetric model} with Bayesian stochastic search variable selection (BSSVS) to estimate significant rates \cite{lemey2009}). Because the posterior set of trees for a single segment has branches labelled with the inferred lineage in the remaining 7 segments, we can detect inter-lineage reassortments between pairs of segments by observing trait transitions, i.e. Yam to Vic or Vic to Yam (see Figure \ref{methodFig}). 
In addition, by reconstructing the ancestral state of all other genomic segments jointly we can infer co-reassortment events when more than one trait transition occurs on the same node in a tree.

@@

\subsection*{Confirming findings with a larger dataset through lineage assignment}
The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences.
Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.
